---
title: "Peer Project 1"
author: "Haley"
date: "6/21/2020"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lattice)
options(scipen = 999)
```

#Loading/Preprocessing Data
Note, include na.strings and change date to correct format
```{r loadprocess}
data <- as.data.frame(read.csv("activity.csv", header = T, sep = ",", na.strings = "NA"))
data$date <- as.Date(data$date)
```

# Analysis
### What is the mean total number of steps taken per day? 
1. Calculate the total number of steps taken per day
2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
3. Calculate and report the mean and median of the total number of steps taken per day
```{r q1}
data_nona <- na.omit(data)
#aggregate(data_nona$steps, by = list(date = data_nona$date), FUN = mean)
tot_steps <- aggregate(data_nona$steps, by = list(date = data_nona$date), FUN = sum)
hist(tot_steps$x, pch = 20, xlab = "Total Steps", ylab = "Frequency", main = "Frequency of Daily Steps")

p50 <- mean(tot_steps$x)
med <- median(tot_steps$x)
```

The mean and median total number of steps taken per day are `r p50` and `r med`, repsectively. 

### What is the average daily activity pattern?
1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

``` {r q2}
avg_steps <- aggregate(data_nona$steps, by = list(date = data_nona$date, interv = data_nona$interval),
     FUN = mean)
plot(avg_steps$interv, avg_steps$x, type = "l", main = "Average # of Steps Taken, Avg. Across All Days", 
     ylab = "Avg. Steps", xlab = "5-Minute interval")
max_avg <- max(avg_steps$x)
max_avg <- avg_steps[which(avg_steps$x == max_avg ), ]
```
The 5-minute interval, on average across all the days in the dataset, containing the maximum number
of steps is `r max_avg$interv`. 

### Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with \color{red}{\verb|NA|}NAs)
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r q3}
na_rows <- nrow(data[which(is.na(data$steps) == T), ])

#Fill in missing values: method = fill in NAs with mean (chosen because mean/median were so similar)
data_fill_na <- data
data_fill_na[is.na(data_fill_na$steps), 1] <- p50

#Calculate mean/median of total number of steps taken per day and make histogram 
tot_steps2 <- aggregate(data_fill_na$steps, by = list(date = data_fill_na$date), FUN = sum)
hist(tot_steps2$x, pch = 20, xlab = "Total Steps", ylab = "Frequency", main = "Frequency of Daily Steps")

p50_2 <- mean(tot_steps2$x)
med_2 <- median(tot_steps2$x)
```
There are `r na_rows` with missing values. These values differ quite a lot. The mean is much higher, 
but the median is not too different. Neglecting the NA values generates a bimodal distribution with 
a huge variance. 


### Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
Make a panel plot containing a time series plot (i.e. \color{red}{\verb|type = "l"|}type = "l") of the 2. 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r q4}
data_fill_na$dow <- weekdays(data_fill_na$date)
data_fill_na$week <- 0
data_fill_na$week[which(data_fill_na$dow %in% c("Saturday", "Sunday"))] <- 1
data_fill_na$week <- factor(data_fill_na$week, labels = c("Weekday", "Weekend"))

avg_steps2 <- aggregate(data_fill_na$steps, by = list(interv = data_fill_na$interval, wk = 
              data_fill_na$week),FUN = mean)

avg_steps2 <- as.data.frame(avg_steps2)
as2_weekday <- avg_steps2[which(avg_steps2$wk == "Weekday"), ]
as2_weekend <- avg_steps2[which(avg_steps2$wk == "Weekend"), ]

par(mfrow = c(2,1))
plot(as2_weekday$interv, as2_weekday$x, type = "l", main = "Weekday", ylab = "Avg. Steps", xlab = "")
plot(as2_weekend$interv, as2_weekend$x, type = "l", main = "Weekend", ylab = "Avg. Steps", xlab = "5-Minute interval")

```