---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r loadprocess}
library(tidyverse)
library(lattice)
options(scipen = 999)
data <- as.data.frame(read.csv("activity.csv", header = T, sep = ",", na.strings = "NA"))
data$date <- as.Date(data$date)
```

## What is mean total number of steps taken per day?
```{r q1}
data_nona <- na.omit(data)
#aggregate(data_nona$steps, by = list(date = data_nona$date), FUN = mean)
tot_steps <- aggregate(data_nona$steps, by = list(date = data_nona$date), FUN = sum)
hist(tot_steps$x, pch = 20, xlab = "Total Steps", ylab = "Frequency", main = "Frequency of Daily Steps")

p50 <- mean(tot_steps$x)
med <- median(tot_steps$x)
```
The mean and median total number of steps taken per day are `r p50` and `r med`, repsectively. 

## What is the average daily activity pattern?
``` {r q2}
avg_steps <- aggregate(data_nona$steps, by = list(date = data_nona$date, interv = data_nona$interval),
     FUN = mean)
plot(avg_steps$interv, avg_steps$x, type = "l", main = "Average # of Steps Taken, Avg. Across All Days", 
     ylab = "Avg. Steps", xlab = "5-Minute interval")
max_avg <- max(avg_steps$x)
max_avg <- avg_steps[which(avg_steps$x == max_avg ), ]
```
The 5-minute interval, on average across all the days in the dataset, containing the maximum number
of steps is `r max_avg$interv`. 


## Imputing missing values
```{r q3}
na_rows <- nrow(data[which(is.na(data$steps) == T), ])

#Fill in missing values: method = fill in NAs with mean (chosen because mean/median were so similar)
data_fill_na <- data
data_fill_na[is.na(data_fill_na$steps), 1] <- p50

#Calculate mean/median of total number of steps taken per day and make histogram 
tot_steps2 <- aggregate(data_fill_na$steps, by = list(date = data_fill_na$date), FUN = sum)
hist(tot_steps2$x, pch = 20, xlab = "Total Steps", ylab = "Frequency", main = "Frequency of Daily Steps")

p50_2 <- mean(tot_steps2$x)
med_2 <- median(tot_steps2$x)
```
There are `r na_rows` with missing values. These values differ quite a lot. The mean is much higher, 
but the median is not too different. Neglecting the NA values generates a bimodal distribution with 
a huge variance. 


## Are there differences in activity patterns between weekdays and weekends?
```{r q4}
data_fill_na$dow <- weekdays(data_fill_na$date)
data_fill_na$week <- 0
data_fill_na$week[which(data_fill_na$dow %in% c("Saturday", "Sunday"))] <- 1
data_fill_na$week <- factor(data_fill_na$week, labels = c("Weekday", "Weekend"))

avg_steps2 <- aggregate(data_fill_na$steps, by = list(interv = data_fill_na$interval, wk = 
              data_fill_na$week),FUN = mean)

avg_steps2 <- as.data.frame(avg_steps2)
as2_weekday <- avg_steps2[which(avg_steps2$wk == "Weekday"), ]
as2_weekend <- avg_steps2[which(avg_steps2$wk == "Weekend"), ]

par(mfrow = c(2,1))
plot(as2_weekday$interv, as2_weekday$x, type = "l", main = "Weekday", ylab = "Avg. Steps", xlab = "")
plot(as2_weekend$interv, as2_weekend$x, type = "l", main = "Weekend", ylab = "Avg. Steps", xlab = "5-Minute interval")

```
There are differences between average steps during the day for weekdays versus weekends. First, note that we do not see avg. steps increase until later (perhaps respondents are sleeping in). Second, avg. steps taper off later during tegh weekend compared to the weekdays (again, perhaps respondents stay up later). Additionally, note that for the weekend, the average number of steps taken during the 5-minute intervals is more consistent throughout compared to the weekdays. Perhaps this is because during the week, respondents are working and naturally walk less. 